---
title: "Reproducible Research: Peer Assessment 1"
auther: Sahaj G
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data

### Loading and unzippng the data


```{r warning=FALSE}
# graphics library load
library(ggplot2)

# Unzip archive
unzip("activity.zip")

# Read base data into a data frame.
baseData <- read.csv("activity.csv")

# Data Head
head(baseData)

```

### Modifying NA & Date from string class to Date class


```{r}
# change date type to date class 
baseData$date <- as.Date(as.character(baseData$date))
# create NA logical vector
baseDataNA <- is.na(baseData$steps)
# create clean base for later mean calculations
cleanBase <- baseData[!baseDataNA,]
```


## What is mean total number of steps taken per day?

### 1.Calculate the toal number of steps taken per day

```{r warning=FALSE}
# aggregate clean non NA steps per day (SUM)
SummedDataByDay <- aggregate(baseData$steps, by=list(baseData$date), sum)
# adjust column names
names(SummedDataByDay)[1] ="date"
names(SummedDataByDay)[2] ="totalsteps"
# top 15 of Summed Steps by day
head(SummedDataByDay,15)
```

### 2.histogram for the total number of steps taken each day

```{r warning=FALSE}
# Plot using ggplot
ggplot(SummedDataByDay, aes(x = totalsteps)) +
  geom_histogram(fill = "steelblue", binwidth=1000) +
  labs(title = "Total Daily Steps", x = "Steps", y = "Frequency")
```


### 3.Calculating the mean & median of the total number of steps taken per day

```{r}
# Mean of steps taken per day
mean(SummedDataByDay$totalsteps,na.rm=TRUE)
```
```{r}
#median of steps taken per day
median(SummedDataByDay$totalsteps,na.rm=TRUE)
```


## What is the average daily activity pattern?

### 1.A time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

```{r warning=FALSE}
nonNASubset <- baseData[!baseDataNA,]
MeanDataByInterval <- aggregate(nonNASubset$steps, by=list(nonNASubset$interval), mean)
# set the column names
names(MeanDataByInterval)[1] ="interval"
names(MeanDataByInterval)[2] ="steps"

ggplot(MeanDataByInterval, aes(x = interval, y=steps)) +
  labs(title = "Sum of Steps by Interval", x = "5 min interval", y = "steps")+
  geom_line(color="red") 

```


### 2. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

```{r}
maxInterval <- MeanDataByInterval[which.max(MeanDataByInterval$steps),]
maxInterval
```


## Imputing missing values

### 1. Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)

```{r warning=FALSE}
# generate listing of NA's
missingVals <- sum(baseDataNA)
missingVals
```


### 2. Devise a strategy for filling in all of the missing values in the dataset.

#### strategy used : replacing the NA values with the mean interval steps for the corresponding intervals

### 3. Creating a new dataset that is equal to the original dataset but with the missing data filled in.

```{r}
baseData2 <- baseData
missingData <- is.na(baseData2$steps)
meanVals <- tapply(cleanBase$steps, cleanBase$interval, mean, na.rm=TRUE, simplify=TRUE)
baseData2$steps[missingData] <- meanVals[as.character(baseData2$interval[missingData])]
# original missing data count
sum(missingData)
```
```{r}
# count of NA values
sum(is.na(baseData2$steps))
```


### 4a) Histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day.


```{r}
FullSummedDataByDay <- aggregate(baseData2$steps, by=list(baseData2$date), sum)

names(FullSummedDataByDay)[1] ="date"
names(FullSummedDataByDay)[2] ="totalsteps"
head(FullSummedDataByDay,15)
```


```{r}
# Plot using ggplot
ggplot(FullSummedDataByDay, aes(x = totalsteps)) +
  geom_histogram(fill = "steelblue", binwidth=1000) +
  labs(title = "Total Daily Steps", x = "Steps", y = "Frequency")
```


mean of new data


```{r}
mean(FullSummedDataByDay$totalsteps)
```


median of new data 


```{r}
median(FullSummedDataByDay$totalsteps)
```




### 4b) Do these values differ from the estimates from the first part of the assignment?

#### Yes, The mean was unchanged while the meadian increased by 1.19 steps

##### Original mean & median : 10766.19 & 10765
#####      New mean & median : 10766.19 & 10766.19

### 4c) What is the impact of inputing missing data on the estimates of the total daily number of steps?

#### Rplacing the missing data with seems to have pushed the whole data set towards the mean 



## Are there differences in activity patterns between weekdays and weekends?

```{r warning=FALSE}
baseData2$weekday <- weekdays(baseData2$date)
baseData2$weekend <- ifelse (baseData2$weekday == "Saturday" | baseData2$weekday == "Sunday", "Weekend", "Weekday")
#baseData2$weekend <- as.factor(baseData2$weekend)
head(baseData2,5)
```
```{r}
MeanDataWeekendWeekday <- aggregate(baseData2$steps, by=list(baseData2$weekend, baseData2$interval), mean)
names(MeanDataWeekendWeekday)[1] ="weekend"
names(MeanDataWeekendWeekday)[2] ="interval"
names(MeanDataWeekendWeekday)[3] ="steps"

ggplot(MeanDataWeekendWeekday, aes(x = interval, y=steps, color=weekend)) +
  geom_line() +
  facet_grid(weekend ~ .) +
  labs(title = "Mean of Steps by Interval", x = "interval", y = "steps")
```

#### The earlier stages of the day seems to have more number of steps on week days most likely because these are usually the working hours.
#### The weekends on the other hand have the steps fairly spread across the whole day and also seem to have more number of steps than weekdays.
